The Weakly Nonlinear Magnetorotational Instability

### 2016-08-04 paper/widegap_mri_paper.tex
widegap stress figure added to paper

\title{Weakly Nonlinear Analysis of the Standard and Helical Magnetorotational Instabilities in a Taylor Couette Flow}
\author{Clark, S.E.\altaffilmark{1}}
\author{Oishi, J.S. \altaffilmark{2,}\altaffilmark{3}}
%\author{Mac Low, M.-M.\altaffilmark{1,}\altaffilmark{2}}
\altaffiltext{1}{Department of Astronomy, Columbia University, New York, NY} 
\altaffiltext{2}{Department of Astrophysics, American Museum of Natural History, New York, NY}
\altaffiltext{3}{Department of Physics, SUNY Farmingdale}

@@

The magnetorotational instability (MRI) drives angular momentum transport and turbulence in astrophysical disks. Its discovery by \citeauthor{Balbus:1991vs} (\citeyear{Balbus:1991vs}; actually a rediscovery of \citei{Chandrasekhar:1960wh}) was a breakthrough in the longstanding question of how efficient accretion can exist in the universe: that is, how matter collapsing onto a central body is able to coalesce despite the conservation of specific angular momentum. %The ubiquity of disks ...
The ubiquity of astrophysical disks suggests that systems both experience and overcome this centrifugal barrier, and accretion proceeds even in hydrodynamically stable disks. The MRI is excited by weak magnetic fields in differentially rotating fluids, and since its discovery it has been widely invoked to explain accretion in protoplanetary disks () and disks around black holes (), as well as

@@

Because of the MRI's complexity, analytical treatments of the MRI use some combination of convenient approximations, such as ideal MHD, linearized equations, expedient boundary conditions, and simplified geometries. In particular, much of the analytical work on MRI saturation has focused on local approximations, often representing a section of a disk by solving the MHD equations in an isolated region subject to shear periodic boundary conditions (the ``shearing box"). While the MRI is a local instability, certain properties of the local problem are not generic to the global problem. In the shearing box, linear evolution is dominated by a class of MRI mode known as channel modes.  %Channel modes, a class of MRI mode that dominates MRI evolution and saturation in the local approximation, are artificially over-prominent in the shearing box (e.g. \citei{Latter:2015}). 
These are linear modes which also happen to be exact solutions to the \textit{nonlinear} local MRI equations. Runaway growth is avoided in this paradigm by the instability of the channel modes themselves, which are destroyed by parasitic (secondary) instabilities (\citei{Goodman:1994ul}, \citei{Pessah:2010ic}). The growth of parasitic modes provides a saturation avenue for channel mode-dominated flows, yet this is unlikely to be the dominant saturation mechanism in laboratory experiments or astrophysical disks, as channel modes are artificially over-prominent in the shearing box (e.g. \citei{Latter:2015}).

The theory we develop here employs radial boundary conditions, and so MRI modes are global and channel modes are not present. A number of saturation mechanisms have been proposed for the MRI which do not rely on channel modes dominating the flow. The MRI feeds off of the free energy from differential rotation, and so a modification of the background shear may cause saturation (e.g. \citei{Knobloch:2005ba}, \citei{Umurhan:2007hs}). The MRI may transfer its free energy into the magnetic field, and saturate when the field is too strong to be susceptible to the MRI (e.g. \citei{Ebrahimi:2009ey}). The MRI may saturate differently depending on the particular parameter regime under investigation, and so our challenge in not only in identifying possible saturation mechanisms, but in understanding how and when each applies in different astrophysical environments.

@@

@@

>>>>

%\title{Weakly Nonlinear Analysis of the Standard and Helical Magnetorotational Instabilities in a Taylor Couette Flow}
\title{The Weakly Nonlinear Magnetorotational Instability} % maybe this is punchier?
\author{Clark, S.E.\altaffilmark{1}}
\author{Oishi, J.S. \altaffilmark{2,}\altaffilmark{3}}
%\author{Mac Low, M.-M.\altaffilmark{1,}\altaffilmark{2}}
\altaffiltext{1}{Department of Astronomy, Columbia University, New York, NY} 
\altaffiltext{2}{Department of Astrophysics, American Museum of Natural History, New York, NY}
\altaffiltext{3}{Department of Physics, SUNY Farmingdale}

@@

The magnetorotational instability (MRI) drives angular momentum transport and turbulence in astrophysical disks. Its discovery by \citeauthor{Balbus:1991vs} (\citeyear{Balbus:1991vs}; actually a rediscovery of \citei{Chandrasekhar:1960wh}) was a breakthrough in the longstanding question of how efficient accretion can exist in the universe: that is, how matter collapsing onto a central body is able to coalesce despite the conservation of specific angular momentum. %The ubiquity of disks ...
The ubiquity of astrophysical disks suggests that systems both experience and overcome this centrifugal barrier, and accretion proceeds even in hydrodynamically stable disks. The MRI is excited by weak magnetic fields in differentially rotating fluids, and since its discovery it has been widely invoked to explain accretion in protoplanetary disks (see \citei{Armitage:2010} and references therein) and disks around black holes (), as well as

@@

Because of the MRI's complexity, analytical treatments of the MRI use some combination of convenient approximations, such as ideal MHD, linearized equations, expedient boundary conditions, and simplified geometries. In particular, much of the analytical work on MRI saturation has focused on local approximations, often representing a section of a disk by solving the MHD equations in an isolated region subject to shear periodic boundary conditions (the ``shearing box"). The shearing box is an inexpensive computational framework, so is used by many numerical MRI studies as well. However, while the MRI is a local instability, certain properties of the local problem are not generic to the global problem. In the shearing box, linear evolution is dominated by a class of MRI mode known as channel modes.  %Channel modes, a class of MRI mode that dominates MRI evolution and saturation in the local approximation, are artificially over-prominent in the shearing box (e.g. \citei{Latter:2015}). 
These are linear modes which also happen to be exact solutions to the \textit{nonlinear} local MRI equations. Runaway growth is avoided in this paradigm by the instability of the channel modes themselves, which are destroyed by parasitic (secondary) instabilities (\citei{Goodman:1994ul}, \citei{Pessah:2010ic}). The growth of parasitic modes provides a saturation avenue for channel mode-dominated flows, yet this is unlikely to be the dominant saturation mechanism in laboratory experiments or astrophysical disks, as channel modes are artificially over-prominent in the shearing box (e.g. \citei{Latter:2015}).

The theory we develop here employs realistic radial boundary conditions, and so MRI modes are global and channel modes are not present. A number of saturation mechanisms have been proposed for the MRI which do not rely on channel modes dominating the flow. The MRI feeds off of the free energy from differential rotation, and so a modification of the background shear may cause saturation (e.g. \citei{Knobloch:2005ba}, \citei{Umurhan:2007hs}). The MRI may transfer its free energy into the magnetic field, and saturate when the field is too strong to be susceptible to the MRI (e.g. \citei{Ebrahimi:2009ey}). The MRI may saturate differently depending on the particular parameter regime under investigation, and so our challenge is not only in identifying possible saturation mechanisms, but in understanding how and when each applies in different astrophysical environments.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../figures/widegap_rey_max_tot_stresses.png}
\caption{Reynolds, Maxwell, and total stress for the fiducial standard MRI case. }
\end{figure*}

@@

\bibitem[{Armitage(2010)}]{Armitage:2010}
Armitage, P. 2010, ARA\&A, 49, 195
